The unphysicality of Hilbert spaces

### 2024-11-21 Articles/2023-HilbertQuantum/Submissions/Quantum Studies/HilbertQuantum.tex
Updates after Francisco's comments and further proofreading

\begin{abstract}
We show that Hilbert spaces are not suitable to represent quantum states mathematically, in the sense that they require properties that are untenable by physical entities. We first prove that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice-versa, and time evolutions that transform finite expectations to infinite ones in finite time. This means we should be wary of using Hilbert spaces to represent quantum states as they turn a potential infinity into an actual infinity. The main point of the paper, then, is to raise awareness that results that rely on the completeness of Hilbert spaces may not be physically significant. While we do not claim to know what a physically more appropriate closure should be, we note that Schwartz spaces, among other things, guarantee that the expectation of all polynomials of position and momentum are finite, their elements are uniquely identified by these expectations, and they are closed under Fourier transform.
\end{abstract}

@@

In an effort to find a precise physical justification, we realized that no such justification can exist as infinite-dimensional Hilbert spaces always contain elements that cannot have physical counterparts. To put it simply, they must include states that have infinite expectation values for position, momentum, energy and so on. This means that a truly general statement in quantum theory must be valid also on these irrelevant unphysical cases, with the consequence that either physical results are left out, or they are obscured in a complicated patchwork of theorems with different domains of validity. We therefore suspect that some of the mathematical difficulties encountered when trying to give a more precise mathematical foundation to quantum theory stem from this fundamental mismatch.

In this work we briefly present our findings, leaving more mathematical detail to the appendix. We will see that complex inner product spaces are indeed needed, and should be considered physical. We will see that the problem lies in requiring completeness, which should be understood as an unphysical mathematical requirement. We will show that other options exist, offering Schwartz spaces as an example. These guarantee the expectation values of all polynomials of position and momentum to be finite, and they are also closed under Fourier transform.

@@

When we first drafted this paper, we naively thought that the advantage of having a well-posed mathematical structure that represents, and only represents, physically well-defined objects would be self-evident. It does not seem to be the case: a reviewer from a previous journal submission stated that ``the question of whether the limiting states are actually physical is rather pointless;'' conversely, others see no problem in working with quantum field theory path integrals that are not mathematically well-defined. We are therefore in the absurd, from our perspective, position of having to defend that, yes, the math we need to use in physics has be mathematically precise and, yes, the math we need to use in physics has to represent---perhaps idealized---physically realizable objects. The cost of this failure is all around us: an ever-increasing number of interpretations of quantum mechanics, abstract mathematical work whose connection to physics is unclear, theoretical physics more and more disconnected from experimental physics. The reader who, like us, finds this obvious can skip this section.

One problem is that most physicists see mathematics as just a tool for calculation, much like oscilloscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say, ``States are rays in a Hilbert space,'' we are giving a physical specification. We are saying that there are objects in the real world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

@@

While it seems obvious to us, it may be worth stating that \textbf{it is not up to mathematicians to justify the use of a particular mathematical structure in physics}. Ultimately, the justification cannot be solely in terms of mathematical matters, but rather it has to rely on that elusive ``physical intuition.'' In fact, the task is exactly to make that elusive ``physical intuition'' precise enough to turn it into mathematical definitions. This is the job of a physicist. In our specific case, the core argument we will use is that states described by measurable quantities with infinite or ill-defined expectation values are not physically realizable. To us, and to every experimental physicist we talked to, ``It doesn't make sense experimentally,'' and this is the best argument possible to show that something is unphysical. ``It can't be done.'' No argument is better. However, more than one previous reviewer found no problem with infinite (or undefined) expectation values, as they adopt an ``ensemble interpretation,'' and tried to convince us that we should not be concerned either. How can progress on the mathematical foundations of quantum mechanics be made if an argument that is accepted by the vast majority of physicists is rejected when one tries to discuss it in the literature?

If the problem of justifying the use of mathematical structures in physics is, ultimately, a physical problem, it also follows that it cannot be solved simply by a different mathematical approach. For example, rigged Hilbert spaces were developed to give a mathematical foundation to Dirac’s bra-ket formalism\cite{Dirac1947}, and to be able to talk about eigenstates of position and momentum, working mathematically with both continuous and discrete bases on an interchangeable level\cite{Madrid_2005,Celeghini2016}. The question is not whether these can be useful at a mathematical level to solve certain problems. They clearly are. The question at hand is whether, on physical grounds, this is a good idea. Should we really treat continuous and discrete spectra the same, or should we consider that they are, on physical grounds, two rather different things? Should the fact that the energy spectrum is typically discrete for bound states and continuous for unbound states make us wonder if there is something in the differing topologies of the spectra that is physically important? Should we then make the difference more prominent in our thinking, rather than trying to construct mathematical tools that treat them the same?  Similarly, one can take the algebraic route and start with an algebra of observables. This, some argue, makes some mathematical problems easier to solve, which indeed it does. It still begs the question: why is it that, on physical grounds, observables should form a $C^*$-algebra? What, then, is the suitable physically meaningful state space? Moreover, note that in both these frameworks, Hilbert spaces still play a crucial role. If Hilbert spaces are physically problematic, these approaches are still relying on a mathematical structure that is physically problematic.

While the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implications for both the physics and the math. It is often the case that the mathematical detail informs the physics and should not be swept under the rug.  Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second-countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true\cite{aop-book}. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second-countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e.~non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the evidence-based requirement of physics justifies the mathematical requirements.

To summarize, we are not interested here in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. Moreover, the fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. States in physics are the output of a preparation procedure, a characterization that works for both ``pure states'' of mechanics as well as ensembles in statistical mechanics. They are what we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?

@@

For example, suppose we want to define volumes in three-dimensional Euclidean space $\mathbb{R}^3$. Formally, we want to find a measure $\mu(U)$ that returns the volume of a region $U$. Naively, we could take a region to be any subset of $\mathbb{R}^3$. However, this leads to the Banach–Tarski paradox,\cite{tao2021introduction} which states that we can take a unit ball, cut it into five pieces, and reassemble those pieces into two balls each of the same volume as the original. Clearly, this does not make physical sense. However, we get a consistent definition if we restrict regions to be Borel sets, instead of arbitrary subsets. Therefore, bigger is not always better in mathematics:\footnote{In another work\cite{aop-book} we show that the Borel sets are exactly those sets that can be associated with experimental procedures. We have often found that physical requirements naturally map to well-behaved spaces.} we need to look more closely at what the math dragged in.

Unitary transformations play an important role in the theory of Hilbert spaces for at least two reasons. The first is that two Hilbert spaces that differ by a unitary transformation are mathematically equivalent. For example, changes of basis and changes of variables are unitary transformations. This is already a problem: the space of wave-functions for $n$ particles is the Hilbert space $L^2(\mathbb{R}^{3n})$. For any value of $n$, these infinite-dimensional spaces are separable, meaning that they all have a countable basis, and so are unitarily equivalent. That is, the state space for one particle is the same as the state space for, say, a billion particles. We can therefore theoretically construct a quantum channel that encodes the information of a billion particles into a single one. Note that this is not possible in classical mechanics: the state space for $n$ classical particles is $\mathbb{R}^{6n}$. For different $n$, these spaces are not even topologically equivalent.

The second reason for the importance of unitary transformation is that they represent possible time evolution for an isolated system. The issue is that unitary transformations in a Hilbert space can map states with finite expectation values to those with infinite ones, leading to obvious problems.

@@

Moreover, we can create the following family of transformations
\begin{equation}
z(x, t) = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
This is clearly continuous in $t$ and, for $t=0$, we have the identity transformation, and therefore, mathematically, there is no reason why this wouldn't be a valid time evolution. What this does is stretch back and forth the wavefunction, such that the state keeps oscillating between the two functions above, making the expectation value oscillate from finite to infinite in finite time. This, again, is physically untenable.

Again, these issues often do not arise when performing actual calculations because physics makes us work in the corner of the structure where things are reasonable. The issue is that when one tries to prove theorems or general results on quantum theory, one is currently forced to consider the entire Hilbert space. It is unreasonable to expect all theorems valid in all physical cases to have valid generalizations for unphysical ones. We are likely missing out on useful results by not considering what is provable within the subset of physical cases. We, therefore, wonder whether many of the technical issues we have in quantum theory (e.g. the non-existence of a valid measure for path integrals\cite{glimm_quantum_1987}, inequivalent representations for interacting theories\cite{earman_haags_2006}, the non-existence of the categorical tensor product for Hilbert spaces\cite{garrett_tensor_2010, sorkin_inside_2022}, etc.)  are simply a consequence of using unphysical objects at a fundamental level.

@@

While we do need infinite-dimensional spaces to handle unbound quantities, such as the number of particles, or continuous quantities, such as position, it seems that Cauchy-completeness is not the proper characterization. While finding the proper one goes beyond the scope of this work, we want to show that fruitful lines of inquiry do exist.

A reasonable alternative would be to require the expectation of all polynomials of position and momentum to be finite. In quantum mechanics, this means requiring all expectations of $\frac{1}{2}(X^nP^m + P^mX^n)$ to be finite. This restricts our space to those functions that are infinitely smooth (i.e. all derivatives exist) and decrease very rapidly as $x$ goes to infinity: these are the Schwartz functions.

@@

This appendix presents the same arguments in the article's main body, but expanded with full mathematical detail. The figure and the derivations in the Appendix are courtesy of the Assumptions of Physics project. \cite{aop-book}

@@

If $\theta_{\psi\phi}$ is the angle between $\psi$ and $\phi$ on the Bloch sphere, we have
\begin{equation}
	|\langle \psi | \phi \rangle |^2 = \cos^2 \frac{\theta_{\psi\phi}}{2}.
\end{equation}
The angle is divided by two because the angle on the Bloch sphere (i.e. in physical space) is double the angle in the Hilbert space. For example, the angle between $+$ and $-$ on the Bloch sphere is $\pi$, but the inner product between them is zero (i.e. opposite directions in physical space correspond to orthogonal states) and therefore the angle in the Hilbert space is $\pi/2$.

@@

To understand the physical significance, consider the position operator $X : D(X) \to \mathcal{H}$. If we take a normalized vector $\psi \in \mathcal{H}$, the norm $||X\psi|| = \<X\psi|X\psi\> = E[X^2|\psi]$ is equal to the expectation of the position squared. The operator is unbounded because the position can be arbitrarily large. We can therefore construct a Cauchy sequence $\{\psi_i\}$ such that $\lim\limits_{i \to \infty}||X\psi|| = +\infty$, that is, we can take a sequence of vectors for which the expectation of position is larger and larger. Because of completeness, the limit of the Cauchy sequence will be in the Hilbert space. However, because the expectation diverges, the limit cannot be in the domain of $X$. Cauchy completeness forces us to include ``states'' for which the position is ill-defined.

@@

More concretely, a change of variable $x$ to $y$ induces a unitary operator as it is just a change of representation. The transformation rules for wavefunctions simply follow by the fact that $\psi^\dagger(x) \psi(x) = \rho(x)$ gives us a probability density over $x$. Therefore we must have
\begin{equation}
	\begin{aligned}
		\rho(x) dx &= \rho(y) dy \\
		\rho(y) &= \rho(x) \frac{dx}{dy} \\
		\psi^\dagger(y) \psi(y) &= \psi^\dagger(x) \psi(x) \frac{dx}{dy}.
	\end{aligned}
\end{equation}
Therefore we can set 
\begin{equation}
	\begin{aligned}
		\psi(y) &= \psi(x(y)) \sqrt{\frac{dx}{dy}}.
	\end{aligned}
\end{equation}
The above transformation is unitary.

@@

The issue is that the position operator $X$ and the position operator $Y$ are different operators, and the domain of $X$ will be in general different from the domain of $Y$. Let $y=y(x)$ be our change of variable such that $y(0) = 0$. If $\psi(x)$ is a wavefunction expressed over $x$, and $\phi(y)$ represents the same state as a wavefunction over $y$, we have the following relationship:
\begin{equation}
\int_{0}^{y(x)} \phi^\dagger(y) \phi(y) dy = \int_{0}^{x} \psi^\dagger(x) \psi(x) dx
\end{equation}
To find a coordinate transformation that takes finite expectations to infinite expectations, we set the initial and final wavefunctions as follows:
\begin{align}
\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
\phi(y) &= \frac{1}{\sqrt{\pi(y^2 + 1)}}.
\end{align}
The first distribution is a Gaussian; therefore, all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all moments above the first. We find
\begin{equation}
\begin{aligned}
\int_{0}^{y(x)} \frac{1}{\pi(\hat{y}^2 + 1)} d\hat{y} &= \int_{0}^{x} \frac{e^{-\hat{x}^2}}{\sqrt{\pi}} d\hat{x} \\
\frac{\tan^{-1}(y(x))}{\pi} &= \frac{\erf(x)}{2} \\
y(x) &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
\end{aligned}
\end{equation}
The graph of the above function increases so fast that, when plotted, it looks like it has a vertical asymptote. Nonetheless, it is an invertible function over the whole domain and will induce a unitary transformation on the Hilbert space.\footnote{Note that we could implement a similar coordinate transformation in classical mechanics, which would also change finite moments of a distribution into infinite ones. It may be that, like differentiable manifolds are constrained to only differentiable coordinate transformations, we need ``statistical manifolds'' that are constrained to coordinate transformations that preserve the finiteness of moments of distributions. This is a topic that requires further analysis.}

Unitary operators are also used in quantum theory to represent time evolution. Typically, we have a family of unitary operators $\{U_t\}_{t \in \mathbb{R}}$ that is strongly continuous (i.e. $\lim_{t \to t_0} U_t = U_{t_0}$) and respect composition in time (i.e. $U_t U_s = U_{t+s}$). We can turn the above transformation into an evolution by setting
\begin{equation}
z(x, t) = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
The idea is that the evolution will simply stretch and shrink the wavefunction intermittently, potentially oscillating expectation values from finite to infinite and vice versa in finite time. Clearly, this is not a physically meaningful time evolution, yet it will correspond to a strongly continuous one-parameter unitary group and, by Stone's theorem, will even admit a Hamiltonian.

@@

Some physical systems are described by unbounded discrete quantities (e.g. number of particles) or continuous quantities (e.g. position), which require infinite-dimensional spaces. Infinite-dimensional complex inner product spaces, then, are needed to represent quantum systems.

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true for distributions in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of distributions with compact support on phase space gives us finite expectation value for all functions of positions and momentum.}

>>>>

\begin{abstract}
We argue that Hilbert spaces are not suitable to represent quantum states mathematically, in the sense that they require properties that are untenable by physical entities. We first demonstrate that the requirements posited by complex inner product spaces are physically justified. We then show that completeness in the infinite-dimensional case requires the inclusion of states with infinite expectations, coordinate transformations that take finite expectations to infinite ones and vice versa, and time evolutions that transform finite expectations to infinite ones in finite time. This means we should be wary of using Hilbert spaces to represent quantum states as they turn a potential infinity into an actual infinity. The main point of the paper, then, is to raise awareness that results that rely on the completeness of Hilbert spaces may not be physically significant. While we do not claim to know what a physically more appropriate closure should be, we note that Schwartz spaces, among other things, guarantee that the expectation of all polynomials of position and momentum are finite, their elements are uniquely identified by these expectations, and they are closed under Fourier transform.

\end{abstract}

@@

In an effort to find a precise physical justification, we realized that no such justification can exist as infinite-dimensional Hilbert spaces always contain elements that cannot have physical counterparts. To put it simply, they must include states that have infinite expectation values for position, momentum, energy and so on. This means that a truly general statement in quantum theory must be valid also on these irrelevant unphysical cases, with the consequence that either physical results are left out, or they are obscured in a complicated patchwork of theorems with different domains of validity. We therefore suspect that some of the mathematical difficulties encountered when trying to give a more precise mathematical foundation to quantum theory stem from this fundamental mismatch, and we argue more generally that mathematical spaces used in physics should already come equipped with the proper structure that excludes physically pathological behavior.

In this work we briefly present our findings, leaving more mathematical detail to the appendix. We will see that complex inner product spaces are indeed needed, and should be considered physical. We will see that the problem lies in requiring completeness, which should be understood as an unphysical mathematical requirement. We will show that other potential options exist, offering Schwartz spaces as an example. These guarantee the expectation values of all polynomials of position and momentum to be finite, and they are also closed under Fourier transform.

@@

When we first drafted this paper, we naively thought that the advantage of having a well-posed mathematical structure that represents, and only represents, physically well-defined objects would be self-evident. It does not seem to be the case: a reviewer from a previous journal submission stated that ``the question of whether the limiting states are actually physical is rather pointless;'' conversely, others see no problem in working with quantum field theory path integrals that are not mathematically well-defined. We are therefore in the absurd, from our perspective, position of having to defend that, yes, the math we need to use in physics has to be mathematically precise and, yes, the math we need to use in physics has to represent---perhaps idealized---physically realizable objects. The cost of not requiring a clear and tight connection between the math and the physics is all around us: an ever-increasing number of interpretations of quantum mechanics, abstract mathematical work whose connection to physics is unclear, theoretical physics more and more disconnected from experimental physics. The reader who finds this obvious can skip this section.

One problem is that most physicists see mathematics as just a tool for calculation, much like oscilloscopes or telescopes are tools to gather data. As long as they do what they are supposed to do, the details are irrelevant because the physics lies somewhere else. If they do not work correctly, it is just a detail for the expert to fix. But mathematics plays a much more fundamental role in physics: it is the language in which we specify our models. When we say, ``States are rays in a Hilbert space,'' we are giving a physical specification. We are saying that there are objects in the physical world that can be modeled with a particular mathematical structure. If Hilbert spaces do not work to model physical states, it is not simply a math problem: the model is wrong, which is a physics problem. This is the question we are examining in this paper: does the math we use in quantum mechanics model the physical world correctly? Is our physical problem well specified?

@@

While it seems obvious to us, it may be worth stating that \textbf{it is not up to mathematicians to justify the use of a particular mathematical structure in physics}. Ultimately, the justification cannot be solely in terms of mathematical matters, but rather it has to rely on that elusive ``physical intuition.'' In fact, the task is exactly to make that elusive ``physical intuition'' precise enough to turn it into mathematical definitions. This is the job of a physicist. In our specific case, the core argument we will use is that states described by measurable quantities with infinite or ill-defined expectation values are not physically realizable. To us, and to every experimental physicist we talked to, ``It doesn't make sense experimentally,'' and this is the best argument possible to show that something is unphysical. ``It can't be done, not even in principle.''

If the problem of justifying the use of mathematical structures in physics is, ultimately, a physical problem, it also follows that it cannot be solved simply by a different mathematical approach. For example, rigged Hilbert spaces were developed to give a mathematical foundation to Dirac’s bra-ket formalism\cite{Dirac1947}, and to be able to talk about eigenstates of position and momentum, working mathematically with both continuous and discrete bases on an interchangeable level\cite{Madrid_2005,Celeghini2016}. The question is not whether these can be useful at a mathematical level to solve certain problems. They clearly are. The question at hand is whether, on physical grounds, this is a judicious idea. Should we really treat continuous and discrete spectra the same, or should we consider that they are, on physical grounds, two rather different things? Should the fact that the energy spectrum is typically discrete for bound states and continuous for unbound states make us wonder if there is something in the differing topologies of the spectra that is physically important? Should we then make the difference more prominent in our thinking, rather than trying to construct mathematical tools that treat them the same way?  Similarly, one can take the algebraic route and start with an algebra of observables. This, some argue, makes some mathematical problems easier to solve, which indeed it does. It still begs the question: why is it that, on physical grounds, observables should form a $C^*$-algebra? What, then, is the suitable physically meaningful state space? One of the ``successes'' of the algebraic approach was to construct a Hilbert space from an abstract $C^*$-algebra via the so-called GNS construction. Thus in both these frameworks, rigged Hilbert spaces and the $C^*$-algebra approach, Hilbert spaces still play a crucial role. If Hilbert spaces are physically problematic, these approaches are still relying on a mathematical structure that is physically problematic.  Looking to a completely different field, within the hierarchy of safety controls, elimination, substitution, and engineering controls are strongly preferred over administrative controls to manage hazards---putting up a fence at the edge of a cliff is much safer than simply putting a sign saying, ``Beware of cliff.''  By using Hilbert spaces for physical calculations in quantum mechanics, the community is effectively relying on ``administrative controls'' to avoid incorrect results, i.e.~considering if a mathematically allowed calculation makes physical sense and discarding it if not.  For physics problems in realms close to everyday human experience, our physical intuition is very reliable.  However, for problems within e.g.~quantum physics, quantum field theory, or cosmology, our physical intuition is much less powerful, and in fact these areas of physics rely much more heavily on the mathematics.  Our paper is effectively advocating that we try to implement ``engineering controls'' to ensure the soundness of calculations in quantum mechanics---if we can find a suitable mathematical space that does not allow unphysical objects in the first place, then there is no risk of accidentally failing to discard unphysical results.

While the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implications for both the physics and the math. It is often the case that the mathematical detail informs the physics and should not be swept under the rug.  Let us give an example. In a previous work some of us found that a set of physically distinguishable objects must be a $T_0$ second-countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true\cite{aop-book}. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second-countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e.~non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the evidence-based requirement of physics justifies the mathematical requirements.

To summarize, we are not interested here in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. Moreover, the fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. States in physics are the output of a preparation procedure, a characterization that works for both ``pure states'' of mechanics as well as ensembles in statistical mechanics. They are what we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?  
%We note that von Neumann himself placed very high value on conceptual clarity in physical theories.\cite{Redei2006}

@@

For example, suppose we want to define volumes in three-dimensional Euclidean space $\mathbb{R}^3$. Formally, we want to find a measure $\mu(U)$ that returns the volume of a region $U$. Naively, we could take a region to be any subset of $\mathbb{R}^3$. However, this leads to the Banach–Tarski paradox,\cite{tao2021introduction} which states that we can take a unit ball, cut it into five pieces, and reassemble those pieces into two balls each of the same volume as the original. Clearly, this does not make physical sense. However, we get a consistent definition if we restrict regions to be Borel sets, instead of arbitrary subsets. Therefore, bigger is not always better in mathematics:\footnote{In another work\cite{aop-book} some of us show that the Borel sets are exactly those sets that can be associated with experimental procedures. We have often found that physical requirements naturally map to well-behaved spaces.} we need to look more closely at what the math dragged in.

Unitary transformations play an important role in the theory of Hilbert spaces for at least two reasons. The first is that two Hilbert spaces that differ by a unitary transformation are mathematically equivalent. For example, changes of basis and changes of variables are unitary transformations. This is already a problem for physics: the space of wave functions for $n$ particles is the Hilbert space $L^2(\mathbb{R}^{3n})$. For any value of $n$, these infinite-dimensional spaces are separable, meaning that they all have a countable basis, and so are unitarily equivalent. That is, the state space for one particle is the same as the state space for, say, a billion particles. We can therefore theoretically construct a quantum channel that encodes the information of a billion particles into a single one. Note that this is not possible in classical mechanics: the state space for $n$ classical particles is $\mathbb{R}^{6n}$. For different $n$, these spaces are not even topologically equivalent.

The second reason for the importance of unitary transformations is that they represent possible time evolution for an isolated system. The issue is that unitary transformations in a Hilbert space can map states with finite expectation values to those with infinite ones, leading to obvious problems.

@@

Moreover, we can create the following family of transformations
\begin{equation}
z(x, t) = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
This is clearly continuous in $t$ and, for $t=0$, we have the identity transformation, and therefore, mathematically, there is no reason why this would not be a valid time evolution. What this does is stretch back and forth the wave function, such that the state keeps oscillating between the two functions above, making the expectation value oscillate from finite to infinite in finite time. This, again, is physically untenable.

Again, these issues often do not arise when performing actual calculations because physics makes us work in the corner of the structure where things are reasonable. The issue is that when one tries to prove theorems or general results on quantum theory, one is currently forced to consider the entire Hilbert space. It is unreasonable to expect all theorems valid in all physical cases to have valid generalizations for unphysical ones. We are likely missing out on useful results by not considering what is provable within the subset of physical cases. We, therefore, wonder whether many of the technical issues we have in quantum theory (e.g.~the non-existence of a valid measure for path integrals\cite{glimm_quantum_1987}, inequivalent representations for interacting theories\cite{earman_haags_2006}, the non-existence of the categorical tensor product for Hilbert spaces\cite{garrett_tensor_2010, sorkin_inside_2022}, etc.)  are simply a consequence of using unphysical objects at a fundamental level.

@@

While we do need infinite-dimensional spaces to handle unbound quantities, such as the number of particles, or continuous quantities, such as position, it seems that completeness is not the proper characterization. While finding the proper one goes beyond the scope of this work, we want to show that fruitful lines of inquiry do exist.

A reasonable alternative would be to require the expectation of all polynomials of position and momentum to be finite. In quantum mechanics, this means requiring all expectations of $\frac{1}{2}(X^nP^m + P^mX^n)$ to be finite. This restricts our space to those functions that are infinitely smooth (i.e.~all derivatives exist) and decrease very rapidly as $x$ goes to infinity: these are the Schwartz functions.

@@

This appendix presents the same arguments in the article's main body, but expanded with full mathematical detail. The figure and the derivations in the Appendix are courtesy of the Assumptions of Physics project.\cite{aop-book}

@@

If $\theta_{\psi\phi}$ is the angle between $\psi$ and $\phi$ on the Bloch sphere, we have
\begin{equation}
	|\langle \psi | \phi \rangle |^2 = \cos^2 \frac{\theta_{\psi\phi}}{2}.
\end{equation}
The angle is divided by two because the angle on the Bloch sphere (i.e.~in physical space) is double the angle in the Hilbert space. For example, the angle between $+$ and $-$ on the Bloch sphere is $\pi$, but the inner product between them is zero (i.e.~opposite directions in physical space correspond to orthogonal states) and therefore the angle in the Hilbert space is $\pi/2$.

@@

To understand the physical significance, consider the position operator $X : D(X) \to \mathcal{H}$. If we take a normalized vector $\psi \in \mathcal{H}$, the norm $||X\psi|| = \<X\psi|X\psi\> = E[X^2|\psi]$ is equal to the expectation of the position squared. The operator is unbounded because the position can be arbitrarily large. We can therefore construct a Cauchy sequence $\{\psi_i\}$ such that $\lim\limits_{i \to \infty}||X\psi|| = +\infty$, that is, we can take a sequence of vectors for which the expectation of position is larger and larger. Because of completeness, the limit of the Cauchy sequence will be in the Hilbert space. However, because the expectation diverges, the limit cannot be in the domain of $X$. Completeness forces us to include ``states'' for which the position is ill-defined.

@@

More concretely, a change of variable $x$ to $y$ induces a unitary operator as it is just a change of representation. The transformation rules for wave functions simply follow by the fact that $\psi^\dagger(x) \psi(x) = \rho(x)$ gives us a probability density over $x$. Therefore we must have
\begin{equation}
	\begin{aligned}
		\rho(x) dx &= \rho(y) dy \\
		\rho(y) &= \rho(x) \frac{dx}{dy} \\
		\psi^\dagger(y) \psi(y) &= \psi^\dagger(x) \psi(x) \frac{dx}{dy}.
	\end{aligned}
\end{equation}
Therefore we can set 
\begin{equation}
	\begin{aligned}
		\psi(y) &= \psi(x(y)) \sqrt{\frac{dx}{dy}}.
	\end{aligned}
\end{equation}
The above transformation is unitary.

@@

The issue is that the position operator $X$ and the position operator $Y$ are different operators, and the domain of $X$ will be in general different from the domain of $Y$. Let $y=y(x)$ be our change of variable such that $y(0) = 0$. If $\psi(x)$ is a wave function expressed over $x$, and $\phi(y)$ represents the same state as a wave function over $y$, we have the following relationship:
\begin{equation}
\int_{0}^{y(x)} \phi^\dagger(y) \phi(y) dy = \int_{0}^{x} \psi^\dagger(x) \psi(x) dx
\end{equation}
To find a coordinate transformation that takes finite expectations to infinite expectations, we set the initial and final wave functions as follows:
\begin{align}
\psi(x) &= \sqrt{\frac{e^{-x^2}}{\sqrt{\pi}}} \\
\phi(y) &= \frac{1}{\sqrt{\pi(y^2 + 1)}}.
\end{align}
The first distribution is a Gaussian; therefore, all moments of the distribution, the expectation for $x^n$, are finite. The second goes to infinity like $\frac{1}{x^2}$, so the expectation will diverge for all moments above the first. We find
\begin{equation}
\begin{aligned}
\int_{0}^{y(x)} \frac{1}{\pi(\hat{y}^2 + 1)} d\hat{y} &= \int_{0}^{x} \frac{e^{-\hat{x}^2}}{\sqrt{\pi}} d\hat{x} \\
\frac{\tan^{-1}(y(x))}{\pi} &= \frac{\erf(x)}{2} \\
y(x) &= \tan \left(\frac{\pi}{2}\erf(x)\right). \\
\end{aligned}
\end{equation}
The graph of the above function increases so fast that, when plotted, it looks like it has a vertical asymptote. Nonetheless, it is an invertible function over the whole domain and will induce a unitary transformation on the Hilbert space.\footnote{Note that we could implement a similar coordinate transformation in classical mechanics, which would also change finite moments of a distribution into infinite ones. It may be that, like differentiable manifolds are constrained to only differentiable coordinate transformations, we need ``statistical manifolds'' that are constrained to coordinate transformations that preserve the finiteness of moments of distributions. This is a topic that requires further analysis.}

Unitary operators are also used in quantum theory to represent time evolution. Typically, we have a family of unitary operators $\{U_t\}_{t \in \mathbb{R}}$ that is strongly continuous (i.e. $\lim_{t \to t_0} U_t = U_{t_0}$) and respect composition in time (i.e. $U_t U_s = U_{t+s}$). We can turn the above transformation into an evolution by setting
\begin{equation}
z(x, t) = \cos(\omega t) x + \sin(\omega t) \tan \left(\frac{\pi}{2}\erf(x)\right).
\end{equation}
The idea is that the evolution will simply stretch and shrink the wave function intermittently, potentially oscillating expectation values from finite to infinite and vice versa in finite time. Clearly, this is not a physically meaningful time evolution, yet it will correspond to a strongly continuous one-parameter unitary group and, by Stone's theorem, will even admit a Hamiltonian.

@@

Some physical systems are described by unbounded discrete quantities (e.g.~number of particles) or continuous quantities (e.g.~position), which require infinite-dimensional spaces. Infinite-dimensional complex inner product spaces, then, are needed to represent quantum systems.

@@

A naive idea would be to require that the expectation value for all possible variables converges. Unfortunately, this doesn't work. Suppose we have a distribution $\rho(x)$ over position $x$. We would require the expectation of any function to be finite, which would include $e^x$, $e^{e^x}$, $e^{e^{e^x}}$, and so on. The only way this can work is if the function has compact support. Physically, this seems too restrictive as it excludes functions like normal distributions. At a closer investigation, the Paley-Wiener theorem\cite{fourierCompact} tells us that the Fourier transform of a function with compact support is an entire function, and an entire function does not have compact support.\cite{markushevich2014entire} Therefore, we cannot have compact support in both position and momentum.\footnote{This is not true for distributions in classical mechanics, as position and momentum are different variables, not linked by the Fourier transform. Therefore the space of distributions with compact support on phase space gives us finite expectation value for all functions of position and momentum.}
